---
title: "Introduction to `acre`"
format: 
  html:
    embed-resources: true
    self-contained-math: true
title-block-banner: "#046A38"
toc: true
toc-depth: 2
execute:
  warning: false
  message: false
---

```{r include = FALSE}
# devtools::load_all()
library(acre)
```

Just placeholder-ish for now.

:::callout-note
If you haven't already, make sure to install the `acre` using `devtools`.

```{R eval=FALSE}
devtools::install_github("b-steve/acre")
```
:::

The basic `acre` workflow is:

1. Prepare survey data.
2. Combine and process the data using `read.acre()`.
3. Inspect the processed data using `plot()`.
4. Fit a model using `fit.acre()`.
5. Examine and visualize the model results.

This vignette demonstrates each step using a small simulated acoustic survey 
included with the package.

## Example data

The `island_data` dataset contains a simulated acoustic survey conducted on 
Whatupuke Island, New Zealand. It is included to demonstrate the main steps in 
an `acre` analysis.

```{r}
data("island_data")

names(island_data)
```

The dataset is a named list containing:

* `captures`: detector-level records of detected calls;
* `traps`: the locations of the acoustic detectors;
* `mask`: the spatial integration mask;
* `covariates`: spatial measurements of elevation and habitat; and
* `boundary`: coordinates describing the island coastline.

The coordinates are represented in a local Cartesian coordinate system and are 
measured in metres.

### Detection data

The detection data contain one row for every occasion on which a detector 
recorded a call.

```{r}
head(island_data$captures)
```

The three columns required by `acre` are:

* `session`: the survey session in which the detection occurred;
* `ID`: an identifier for the detected call; and
* `trap`: the detector that recorded the call.

The example data also contain a column named `bearing`, giving the estimated 
direction from the detector towards the call in radians.

A call detected at multiple detectors therefore appears in multiple rows, with 
the same `session` and `ID` but a different `trap`.

For example, the call below was detected by several detectors:

```{r}
call_counts <- table(island_data$captures$ID)
example_call <- as.numeric(names(call_counts)[which.max(call_counts)])

subset(
  island_data$captures,
  ID == example_call
)
```

### Detector locations

The detector coordinates are stored in a two-column data frame:

```{r}
island_data$traps
```

Each row corresponds to one detector. Detector identifiers in the `trap` column 
of the detection data refer to the corresponding row of this object.

### Integration mask

The integration mask represents the possible locations of detected calls:

```{r}
head(island_data$mask)
dim(island_data$mask)
```

Each mask point has an associated area:

```{r}
attr(island_data$mask, "area")
```

The mask is used to numerically integrate over unknown call locations during 
model fitting. In this example it has already been restricted to the coastline 
of the island.

### Spatial covariates

The example contains one continuous covariate, `elevation`, and one categorical 
covariate, `habitat`:

```{r}
head(island_data$covariates)

summary(island_data$covariates$elevation)
table(island_data$covariates$habitat)
```

The `x` and `y` columns specify where each covariate measurement was taken. 
These locations do not need to coincide with the integration-mask points because 
`read.acre()` can interpolate spatial covariates onto the mask.

## Reading data into `acre`

The function `read.acre()` combines the survey inputs into an `acre_data` object 
that can be visualized and fitted.

```{r}
island_acre <- read.acre(
  captures = island_data$captures,
  traps = island_data$traps,
  mask = island_data$mask,
  loc.cov = island_data$covariates
)
```

Because a custom island-shaped mask is supplied, `read.acre()` uses it directly. 
If no mask is supplied, one can instead be created automatically from the 
detector locations using the `control.mask` argument.

The resulting object contains the processed capture histories, detector 
locations, integration mask, and spatial covariates:

```{r}
class(island_acre)
names(island_acre)
```

## Visualizing data

Before fitting a model, the prepared survey data should be inspected for obvious 
problems. The `plot()` method for an `acre_data` object can display the detector 
array, capture histories, and interpolated spatial covariates.

### Survey layout

The survey layout shows the integration mask and detector locations:

```{r}
#| label: plot-survey
#| fig-cap: "Integration mask and acoustic detector locations on Whatupuke Island."

plot(
  island_acre,
  type = "survey"
)
```

The grey points represent the integration mask, while the red points represent 
acoustic detectors. The integration mask covers the possible locations of 
detected calls and is used for numerical integration during model fitting.

### Capture histories

Individual capture histories can also be visualized. The following code selects 
the call detected by the largest number of detectors:

```{r}
#| label: choose-example-call

call_counts <- table(island_data$captures$ID)
example_call <- as.numeric(
  names(call_counts)[which.max(call_counts)]
)

example_call
```

We can then plot the detectors that recorded this call together with its 
estimated bearings:

```{r}
#| label: plot-capture
#| fig-cap: "Detector-level capture history and estimated bearings for one call."

plot(
  island_acre,
  type = "capt",
  call_id = example_call,
  ask = FALSE
)
```

Crosses indicate detector locations, while circles identify the detectors that 
recorded the selected call. Arrows show the estimated bearing from each active 
detector towards the call.

### Spatial covariates

The spatial covariates supplied to `read.acre()` have been interpolated onto the 
integration mask. They can be inspected using `type = "covariates"`.

```{r}
#| label: plot-elevation
#| fig-cap: "Elevation interpolated onto the integration mask."

plot(
  island_acre,
  type = "covariates",
  select_cov = "elevation",
  ask = FALSE,
  plot.contours = TRUE
)
```

The categorical habitat covariate can be plotted in the same way:

```{r}
#| label: plot-habitat
#| fig-cap: "Habitat type interpolated onto the integration mask."

plot(
  island_acre,
  type = "covariates",
  select_cov = "habitat",
  ask = FALSE
)
```

These plots should be inspected before fitting a model. In particular, they can 
reveal incorrect detector coordinates, poorly chosen mask limits, or implausible 
spatial interpolation.

## Fitting a model

The function `fit.acre()` fits a spatial capture-recapture model to an object 
returned by `read.acre()`.

We begin with a simple model in which call density and detection parameters are 
spatially constant:

```{r}
#| label: fit-homogeneous
#| cache: true

fit_homogeneous <- fit.acre(
  data = island_acre,
  detfn = "hn",
  tracing = FALSE
)
```

The argument `detfn = "hn"` selects the half-normal detection function. Under 
this model, the probability that a call at location $\mathbf{s}$ is detected by 
detector $j$ is

$$
p_j(\mathbf{s}) = g_0
\exp\left(
-\frac{\lVert \mathbf{s}-\mathbf{x}_j\rVert^2}
{2\sigma^2}
\right),
$$

where $g_0$ is the detection probability at zero distance and $\sigma$ controls 
how quickly detection probability declines with distance.

Because the detection data contain estimated bearings, bearing information is 
included automatically during model fitting. The model therefore also estimates 
`kappa`, which controls the precision of the bearing measurements.

Using `summary()` on the fitted object:

```{r}
summary(fit_homogeneous)
```

### Model formulas

Spatial or survey covariates can be included through the `model` argument. This 
must be a named list in which each name identifies a model parameter and each 
value is an R formula.

For example,

```{r}
model = list(
  D = ~ elevation
)
```

specifies that call density varies with elevation. Parameters omitted from the 
list are treated as spatially constant.

The homogeneous model fitted above is therefore equivalent to specifying 
intercept-only formulas:

```{r}
model = list(
  D = ~ 1,
  g0 = ~ 1,
  sigma = ~ 1,
  kappa = ~ 1
)
```

In practice, these formulas do not need to be supplied because intercept-only 
models are the default.

## Examining model results

Model parameters are estimated internally on link scales. The default call to 
`coef()` returns these linked estimates:

```{r}
coef(fit_homogeneous)
```

For interpretation, it is usually more convenient to return parameters on their 
original scales:

```{r}
coef(
  fit_homogeneous,
  types = "fitted"
)
```

Approximate standard errors and confidence intervals can be extracted in the 
same way:

```{r}
stdEr(
  fit_homogeneous,
  types = "fitted"
)

confint(
  fit_homogeneous,
  types = "fitted"
)
```

For this acoustic survey, `D` represents the estimated density of calls rather 
than the density of individual animals. Estimating animal density would 
additionally require information linking calls to individuals or independently 
collected cue-rate data.

### Modelling spatial variation in density

The homogeneous model assumes that expected call density is constant across the 
island. Spatial covariates can be included using the `model` argument.

Here, we allow call density to vary with habitat and elevation:

```{r}
#| label: fit-spatial-density
#| cache: true

fit_spatial <- fit.acre(
  data = island_acre,
  model = list(
    D = ~ habitat + elevation
  ),
  detfn = "hn",
  tracing = FALSE
)
```

Parameters omitted from the `model` list remain constant. Thus, this model 
allows `D` to vary spatially while retaining constant values of `g0`, `sigma`, 
and `kappa`.

The density component is modelled using a log link:

$$
\log D(\mathbf{s}) = \beta_0
+
\beta_{\mathrm{forest}}
I\{\operatorname{habitat}(\mathbf{s})=\mathrm{forest}\}
+
\beta_{\mathrm{elevation}}
\operatorname{elevation}(\mathbf{s}).
$$

The log link ensures that fitted call density is positive throughout the island.

```{r}
summary(fit_spatial)
```

### Interpreting covariate effects

The linked-scale coefficients describe how the log of call density changes with 
each covariate:

```{r}
coef(
  fit_spatial,
  types = "linked",
  pars = "D"
)
```

The habitat coefficient compares forest with the reference habitat level, 
coastal habitat. A positive coefficient indicates higher estimated call density 
in forest.

The elevation coefficient describes the change in log density associated with a 
one-unit increase in elevation. Since elevation is measured in metres, this 
coefficient may be small even when elevation has a meaningful cumulative effect 
across the island.

Approximate uncertainty can be obtained using:

```{r}
confint(
  fit_spatial,
  types = "linked",
  pars = "D"
)
```

### Visualizing estimated density

The fitted spatial model can be used to estimate call density at each 
integration-mask point.

```{r}
#| label: plot-fitted-density
#| fig-cap: "Estimated call-density surface from the spatial model."

plot(
  fit_spatial,
  type = "Dsurf"
)
```

Areas with larger fitted values correspond to locations where the model predicts 
a greater density of calls.

The fitted density surface reflects both habitat and elevation because these 
covariates were included in the model for `D`.

### Comparing models

The homogeneous and spatial models can be compared using Akaike's information 
criterion:

```{r}
AIC(
  fit_homogeneous,
  fit_spatial
)
```

A smaller AIC indicates stronger support relative to the other models 
considered, after accounting for model complexity.

This comparison should be interpreted as evidence about the models fitted here, 
rather than proof that one covariate structure is the true ecological process.

## The complete workflow

The main `acre` workflow can be summarized as:

```{r eval = FALSE}
# Prepare the data
island_acre <- read.acre(
  captures = island_data$captures,
  traps = island_data$traps,
  mask = island_data$mask,
  loc.cov = island_data$covariates
)

# Inspect the survey
plot(island_acre, type = "survey")

# Fit a model
fit <- fit.acre(
  data = island_acre,
  model = list(
    D = ~ habitat + elevation
  ),
  detfn = "hn",
  tracing = FALSE
)

# Examine parameter estimates
coef(fit, types = "Dsurf")
confint(fit, types = "Dsurf")

# Visualize fitted density
plot(fit, type = "D")
```

## Where to go next

This vignette introduced the standard workflow for preparing, visualizing, 
fitting, and summarizing an `acre` model.

Further vignettes describe:

* modelling more complex forms of spatial variation;
* incorporating different types of auxiliary acoustic information; and
* fitting multi-session models.

