---
title: "Multi-session acoustic survey analysis with `acre`"
format: 
  html:
    embed-resources: true
    self-contained-math: true
title-block-banner: "#046A38"
toc: true
toc-depth: 2
execute:
  warning: false
  message: false
---

```{r include=FALSE}
library(lubridate)
devtools::load_all()
```

::: callout-note
If you have not already installed `acre`, it can be installed from GitHub using
`devtools`.

```{r eval=FALSE}
devtools::install_github("b-steve/acre")
```
:::

A multi-session analysis combines data from repeated surveys and estimates a
single model across all sessions. This allows ecological quantities such as
density, and observation-process parameters such as detectability, to vary with
session-level covariates.

The basic workflow is:

1. Prepare the data for each survey session.
2. Combine the sessions using `read.acre()`.
3. Inspect the processed data using `plot()`.
4. Fit a multi-session model using `fit.acre()`.
5. Examine and visualize variation in density and detectability.

This vignette demonstrates the workflow using repeated acoustic surveys of
frogs. The fitted model asks whether frog call density varies through time and
between sites, while allowing detectability to differ between sites.

## Example data

The frog data consist of repeated acoustic survey sessions. For each session,
we have:

- a detection file containing detected frog calls;
- a set of microphone locations;
- a survey date;
- a site identifier;
- a signal-strength cutoff used to remove weak detections.

Each survey is treated as one session in the combined analysis.

```{r eval=FALSE}
data_dir <- "vignette-data/frog-data"

data_files <- list.files(data_dir, pattern = "data_out_of_Access")

survey_info <- data.frame(
  file = data_files,
  date = substr(data_files, 1, 8),
  site = substr(data_files, 13, 13),
  cutoff = c(
    140, 155, 147.5, 157.5, 152.5, 140, 155, 157.5,
    150, 157.5, 155, 145, 155, 140, 155, 157.5,
    155, 157.5, 147.5, 157.5, 162.5, 155, 152.5,
    155, 157.5, 152.5, 145, 155, 140, 147.5, 145
  )
)

survey_length <- 15
survey_length_hours <- survey_length / 60

n_sessions <- nrow(survey_info)

head(survey_info)
```

The survey length is recorded in minutes. When fitting the model, we pass the
survey length in hours.

We also create the session-level covariates used in the model.

```{r eval=FALSE}
survey_date <- ymd(survey_info$date)

t <- as.numeric(
  as.duration(interval(min(survey_date), survey_date)),
  "days"
)

site <- factor(survey_info$site)

session_cov <- data.frame(
  session = seq_len(n_sessions),
  site = site,
  t = t
)

head(session_cov)
```

Here, `t` is the number of days since the first survey. The variable `site` is a
factor identifying the survey site.

## Preparing the survey sessions

Before fitting an acoustic spatial capture-recapture model, the raw detections
need to be converted into capture histories.

For these data, the main preparation steps are:

1. restrict detections to the survey time window;
2. identify which microphone detected each call;
3. group detections that plausibly came from the same calling frog;
4. remove detections below the session-specific signal-strength cutoff;
5. drop signal strength before model fitting;
6. combine all sessions into one multi-session data object.

::: {.callout-note}
## How signal strength is used here

In this workshop, signal strength is used only as a **filtering rule**. We remove
weak detections below the session-specific cutoff, but we do not use signal
strength in the fitted model.

The fitted model uses time-of-arrival information, microphone locations, and the
spatial capture-recapture model structure.
:::

The call-allocation step is deliberately hidden here because it is a
data-preparation detail for this data set, not the focus of the workshop. In a
real analysis, call allocation is an important modelling and data-processing
choice, and should be handled carefully.

```{r include=FALSE, eval=FALSE}
allocate_calls <- function(mics, dets, sound_speed) {
  mics <- as.matrix(mics)

  trap_dists <- distances(mics, mics)
  n_dets <- nrow(dets)

  dist_mat <- detection_dists(trap_dists, dets$trap)
  timediff_mat <- detection_timediffs(dets$toa, dets$trap)
  maxtime_mat <- dist_mat / sound_speed

  match_mat <- timediff_mat <= maxtime_mat

  incomplete_blocks <- find_incomplete_blocks(match_mat)
  n_blocks <- max(incomplete_blocks)

  final_mat <- matrix(FALSE, nrow = n_dets, ncol = n_dets)
  reqss_mat <- dist_mat / timediff_mat

  for (i in seq_len(n_blocks)) {
    block <- match_mat[incomplete_blocks == i, incomplete_blocks == i]
    reqss <- reqss_mat[incomplete_blocks == i, incomplete_blocks == i]

    is_complete <- all(block)

    if (!is_complete) {
      block <- blockify(block, reqss)
    }

    final_mat[incomplete_blocks == i, incomplete_blocks == i] <- block
  }

  find_incomplete_blocks(final_mat)
}

prepare_session <- function(i, survey_info, data_dir, survey_length) {
  traps <- read.csv(
    file.path(data_dir, paste0("Site", survey_info$site[i], "_estlocs.csv")),
    header = TRUE
  )

  dets <- read.csv(
    file.path(data_dir, survey_info$file[i]),
    header = TRUE
  )

  time_range <- c(20 * 60, 20 * 60 + survey_length)

  dets <- dets[
    dets$startSeconds >= time_range[1] &
      dets$startSeconds <= time_range[2],
  ]

  trap_id <- log2(dets$channelMap) + 1

  ids <- allocate_calls(
    mics = traps,
    dets = data.frame(
      trap = trap_id,
      toa = dets$startSeconds
    ),
    sound_speed = 330
  )

  captures <- data.frame(
    session = 1,
    ID = ids,
    trap = trap_id,
    ss = dets$amplitude,
    toa = dets$startSeconds
  )

  captures <- captures[captures$ss >= survey_info$cutoff[i], ]

  captures_toa <- captures[, c("session", "ID", "trap", "toa")]
  captures_toa$session <- i

  list(
    captures = captures_toa,
    traps = traps
  )
}
```

For workshop speed, we use the first 15 sessions. To use all available sessions,
replace `seq_len(15)` with `seq_len(n_sessions)`.

```{r eval=FALSE}
sessions_to_use <- seq_len(15)

sessions <- lapply(
  sessions_to_use,
  prepare_session,
  survey_info = survey_info,
  data_dir = data_dir,
  survey_length = survey_length
)

full_captures <- do.call(
  rbind,
  lapply(sessions, function(x) x$captures)
)

full_traps <- lapply(
  sessions,
  function(x) x$traps
)

session_cov_workshop <- session_cov[sessions_to_use, ]

head(full_captures)
```

## Reading data into `acre`

The prepared capture histories and detector locations can now be combined into
one multi-session `acre_data` object.

```{r eval=FALSE}
full_data <- read.acre(
  captures = full_captures,
  traps = full_traps,
  control.mask = list(buffer = 40, spacing = 1),
  survey.length = survey_length_hours,
  session.cov = session_cov_workshop
)
```

The important pieces are:

- `captures`: the combined capture histories from all sessions;
- `traps`: a list of microphone locations, one element per session;
- `session.cov`: session-level covariates such as site and time;
- `survey.length`: the survey duration in hours.

## Inspecting the processed data

Before fitting a model, the processed data should be checked for obvious
problems. A simple first check is the number of detections and unique calls in
each session.

```{r eval=FALSE}
detections_by_session <- aggregate(
  trap ~ session,
  data = full_captures,
  FUN = length
)

names(detections_by_session)[2] <- "n_detections"

calls_by_session <- aggregate(
  ID ~ session,
  data = full_captures,
  FUN = function(x) length(unique(x))
)

names(calls_by_session)[2] <- "n_calls"

session_summary <- merge(
  session_cov_workshop,
  detections_by_session,
  by = "session"
)

session_summary <- merge(
  session_summary,
  calls_by_session,
  by = "session"
)

session_summary
```

### Survey layout and capture histories

The `plot()` method for `acre` data objects can be used to visualize the survey
layout and the detections.

First, plot the mask and microphone array for one session.

```{r eval=FALSE}
example_session <- 1

plot(
  full_data,
  type = "survey",
  session = example_session
)
```

Next, plot the capture data for the same session.

```{r eval=FALSE}
plot(
  full_data,
  type = "capt",
  session = example_session,
  call_id = 1
)
```

These plots are useful sanity checks before fitting the model. They help confirm
that the mask, microphone array, and detections look sensible.

## Fitting a model

The function `fit.acre()` fits a spatial capture-recapture model to the object
returned by `read.acre()`.

Here, call density is allowed to vary by site and through time, while the
detection parameters are allowed to vary by site.

The model formulas are:

```{r eval=FALSE}
model = list(
  D = ~ site + t + I(t^2),
  lambda0 = ~ site,
  sigma = ~ site
)
```

This separates the ecological component of the model, represented by `D`, from
the observation component, represented by `lambda0` and `sigma`.

```{r cache=TRUE, eval=FALSE}
trend_fit <- fit.acre(
  full_data,
  detfn = "hhn",
  model = list(
    D = ~ site + t + I(t^2),
    lambda0 = ~ site,
    sigma = ~ site
  ),
  tracing = FALSE
)

summary(trend_fit)
```

The density model asks whether frog density differs between sites and changes
through time. The quadratic time effect allows the density trend to bend rather
than forcing a straight-line change.

The detection model allows detectability to differ between sites. This matters
because a lower number of detections at one site might reflect lower density,
lower detectability, or both.

## Examining model results

### Fitted detection functions

We can visualize the fitted detection function. Because `lambda0` and `sigma`
vary by site, we provide site-level covariate values using `newdata`.

```{r eval=FALSE}
site_levels <- levels(session_cov_workshop$site)

detfn_newdata <- data.frame(
  site = factor(site_levels, levels = site_levels)
)

plot(
  trend_fit,
  type = "detfn",
  newdata = detfn_newdata,
  col = seq_along(site_levels),
  lwd = 2
)

legend(
  "topright",
  legend = paste("Site", site_levels),
  col = seq_along(site_levels),
  lwd = 2,
  bty = "n"
)
```

This plot shows how the fitted detection function changes between sites.

### Fitted density surface

We can also plot the fitted density surface for a particular session.

```{r eval=FALSE}
plot(
  trend_fit,
  type = "Dsurf",
  session = example_session,
  plot.contours = TRUE
)
```

This is a session-specific density surface. In this model, the surface is driven
by the fitted density for that session's site and time.

### Estimated call locations

Finally, we can plot the estimated location contours for a detected call.

```{r eval=FALSE}
example_call <- unique(full_captures$ID[
  full_captures$session == example_session
])[1]

example_call
```

```{r eval=FALSE}
plot(
  trend_fit,
  type = "locations",
  session = example_session,
  call_id = example_call,
  filled = FALSE
)
```

Location plots combine the capture history for a call with the fitted detection
model and density surface. They are useful for checking whether the inferred
call location is plausible given the microphones that detected the call.

### Predicted density trends

To visualize the fitted density trends, we create prediction data over the
observed time range for each site.

```{r eval=FALSE}
tt <- seq(
  min(session_cov_workshop$t),
  max(session_cov_workshop$t),
  length.out = 1000
)

pred_data <- do.call(
  rbind,
  lapply(site_levels, function(s) {
    data.frame(
      t = tt,
      site = factor(s, levels = site_levels)
    )
  })
)
```

Predict density from the fitted model.

```{r eval=FALSE}
pred <- predict(trend_fit, newdata = pred_data)

pred_data$D <- pred$D[, 1]
pred_data$log_D <- log(pred_data$D)
```

Plot the fitted density trends by site.

```{r eval=FALSE}
plot(
  NA,
  xlim = range(session_cov_workshop$t),
  ylim = range(pred_data$log_D, na.rm = TRUE),
  xlab = "Days since first survey",
  ylab = "Predicted log density",
  main = "Predicted density trends by site"
)

cols <- seq_along(site_levels)

for (j in seq_along(site_levels)) {
  site_j <- site_levels[j]
  pd <- pred_data[pred_data$site == site_j, ]

  lines(
    pd$t,
    pd$log_D,
    col = cols[j],
    lwd = 2
  )
}

legend(
  "topright",
  legend = paste("Site", site_levels),
  col = cols,
  lwd = 2,
  bty = "n"
)
```

## The complete workflow

The main steps in this multi-session analysis are:

```{r eval=FALSE}
# Prepare each survey session
sessions <- lapply(
  sessions_to_use,
  prepare_session,
  survey_info = survey_info,
  data_dir = data_dir,
  survey_length = survey_length
)

# Combine sessions into one acre_data object
full_data <- read.acre(
  captures = full_captures,
  traps = full_traps,
  control.mask = list(buffer = 40, spacing = 1),
  survey.length = survey_length_hours,
  session.cov = session_cov_workshop
)

# Fit a multi-session model
trend_fit <- fit.acre(
  full_data,
  detfn = "hhn",
  model = list(
    D = ~ site + t + I(t^2),
    lambda0 = ~ site,
    sigma = ~ site
  ),
  tracing = FALSE
)

# Examine fitted values and plots
summary(trend_fit)
plot(trend_fit, type = "detfn", newdata = detfn_newdata)
plot(trend_fit, type = "Dsurf", session = example_session)
```

This vignette has shown how repeated surveys can be combined into a single
analysis in which call density and detectability vary with session-level
covariates. The same framework can be extended by changing the model formulas
or supplying additional spatial and session-level covariates.
